Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices and raw material differences

### 2022-08-06 supp/03-2dgm-sizeclass.Rmd
<edit index>

Two `size classes` (large/small) were used to examine whether the morphology of the elliptical bifaces from the Millsap Cache and Jowell Farm remain morphologically stable, or express morphological variation. Variability in elliptical biface morphology potentially provides evidence associated with differential resharpening trajectories.

>>>>

Two `size classes` (large/small) were used to assess whether the morphology of elliptical bifaces from Millsap Cache and Jowell Farm remain morphologically stable through reduction. Variability in elliptical biface morphology through reduction provides evidence associated with differential resharpening trajectories.

### 2022-08-06 supp/index.Rmd
<edit index>

--- 
title: "Divergent morphology of elliptical bifaces from sites in the American Southeast articulates with raw material differences"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade and/or exchange with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent in this class due to idiosyncratic responses related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential driven by raw material differences. Additionally, size does not differ between raw materials in the large class; however, in the small class, it does. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife. In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more morphologically diverse than those from the Millsap Cache. Discrete raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

@@

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460]. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not linear). The number of flake scars was subsequently divided by the length of the worked edge, with the heaviest worked area identified by the greater value.

@@

  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material/biogeography.
  -   Bifaces in the _large size class_ recovered from the Millsap Cache and Jowell Farm do not differ in shape or size. This suggests consistent morphology in bifaces from the _large size class_, in support of production-based standardization.
  -   Bifaces in the _small size class_ recovered from the Millsap Cache and Jowell Farm **differ significantly** in size. Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch and/or resharpening.
  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at the Jowell Farm site, providing evidence for a more standardized approach to resharpening and/or retouch at the Millsap Cache site. That finding also supports the hypothesis that greater diversity in shape is associated with retouch and/or resharpening, rather than production/manufacture.

>>>>

--- 
title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices and raw material differences"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade and/or exchange with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent due to idiosyncratic responses related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential driven by raw material differences and reduction practices. The analysis of biface morphlogy by size class demonstrates that size does not differ between raw materials in the large class, but does in the small class. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife. In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more morphologically diverse than those from the Millsap Cache. Distinct reduction practices and raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

@@

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460]. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not a linear metric). The number of flake scars was subsequently divided by the length of the worked edge, with the heaviest worked area identified by the greater value.

@@

  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material/biogeography.
  -   Bifaces in the _large size class_ recovered from the Millsap Cache and Jowell Farm do not differ in shape or size. This suggests consistent morphology in bifaces from the _large size class_, in support of production-based standardization.
  -   Bifaces in the _small size class_ recovered from the Millsap Cache and Jowell Farm **differ significantly** in size. Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch and/or resharpening.
  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to resharpening and/or retouch at Millsap Cache. This finding also provides support for the hypothesis that greater shape diversity is associated with retouch and/or resharpening for elliptical bifaces, rather than production/manufacture.
